sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-03-13 doc/paper-jss/inla-comparison.Rmd
Copy edit

In this example, we will fit a presence-absence dynamic species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1).
We will use a binomial distribution and a logit link.

@@

@@

```{r coef-plot, fig.asp=0.4, out.width="85%", echo=FALSE, fig.cap="Fixed effect estimates from all three models with 95 percent CI.", fig.pos='ht', fig.align='center'}
inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>%
  rename(estimate = mean, conf.low = `0.025quant`, conf.high = `0.975quant`) %>%
  mutate(term = unique(sdmTMB_coefs$term), model = "INLA")

@@

We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map occurrence probabilities in each year in space (Fig.\ \ref{fig:maps}).
For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`---the inverse logit here), to produce occurrence probabilities:

@@

```{r maps, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \\pkg{sdmTMB} model.", fig.align='center'}
inverse_link <- m$family$linkinv
ggplot(pred) +
  geom_raster(aes(X, Y, fill = inverse_link(est))) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

@@

```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted occurence probabilities in time and space from the \\pkg{inlabru} model.", fig.align='center'}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))

>>>>

In this example, we will fit a presence-absence (technically, encounter vs. lack of encounter) dynamic species distribution model with fixed-effect intercepts for year and spatiotemporal random fields structured as AR(1).
We will use a binomial distribution and a logit link.

@@

```{r sdmtmb-sanity-comparison, echo=TRUE}
sanity(m)
```

@@

```{r coef-plot, fig.asp=0.25, out.width="85%", echo=FALSE, fig.cap="Fixed effect estimates from all three models with 95 percent CI.", fig.pos='ht', fig.align='center'}
inla_coefs <- m_inla$summary.fixed[, c("0.025quant", "mean", "0.975quant")] %>%
  rename(estimate = mean, conf.low = `0.025quant`, conf.high = `0.975quant`) %>%
  mutate(term = unique(sdmTMB_coefs$term), model = "INLA")

@@

We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map encounter probabilities in each year in space (Fig.\ \ref{fig:maps}).
For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`---the inverse logit here), to produce encounter probabilities:

@@

```{r maps, out.width="5.5in", fig.cap="Predicted encounter probabilities in time and space from the \\pkg{sdmTMB} model.", fig.align='center'}
inverse_link <- m$family$linkinv
ggplot(pred) +
  geom_raster(aes(X, Y, fill = inverse_link(est))) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed()
```

@@

```{r inlabru-pred, out.width="5.5in", fig.cap="Predicted encounter probabilities in time and space from the \\pkg{inlabru} model.", fig.align='center'}
nd_bru <- nd
nd_bru <- dplyr::left_join(nd_bru, yr_lu, by = "year")
sp::coordinates(nd_bru) <- c("X", "Y")
nd_bru <- as(nd_bru, "SpatialPixelsDataFrame")
pred_bru <- predict(m_bru, data = nd_bru, formula = ~ plogis(f + fac))

### 2023-03-13 doc/paper-jss/owls.Rmd
Copy edit

First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations (Fig.\ \ref{fig:mesh-owls}). The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`). This mesh uses a cutoff distance of 1.5 units (or 150 km).

@@

The result is a mesh with `r mesh$mesh$n` (`mesh$mesh$n`) knots.

@@

We can more formally test this with functions from the \pkg{DHARMa} package.

@@

To produce maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an \pkg{sf} object so that the plot axes will be in latitude and longitude.

@@

Next we retrieve continental coastlines from the package \pkg{rnaturalearth} and outlines of the great lakes from <http://www.naturalearthdata.com/> and transform them to the correct projection.

>>>>

First, we construct a mesh to approximate the spatial process, which is used in the SPDE calculations. The `make_mesh()` function accepts the data coordinates and the minimum allowed distance between vertices (`cutoff`) or approximate number of knots in the mesh (`n_knots`). This mesh uses a cutoff distance of 1.5 units (or 150 km) (Fig.\ \ref{fig:mesh-owls}).

@@

The result is a mesh with `r mesh$mesh$n` (`mesh$mesh$n`) vertices.

@@

We can more formally test this with functions from the \pkg{DHARMa} [@dharma] package.

@@

To produce maps of these predictions, we back transform the spatial coordinates into the same units as our original projection and then convert back into an \pkg{sf} [@pebesma2018] object so that the plot axes will be in latitude and longitude.

@@

Next we retrieve continental coastlines from the package \pkg{rnaturalearth} [@rnaturalearth] and outlines of the great lakes from <http://www.naturalearthdata.com/> and transform them to the correct projection.

### 2023-03-13 doc/paper-jss/paper-jss.Rmd
Copy edit

\pkg{sdmTMB} would not be possible without the \pkg{TMB} [@kristensen2016] and \proglang{R}-\pkg{INLA} [@rue2009; @lindgren2011; @lindgren2015] \proglang{R} packages.
\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages.
Smoother support was possible thanks to \pkg{mgcv} [@wood2017a].
We thank the authors of all these packages.
We thank S. Kotwicki, M. Lindmark, M. Martin, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
Christmas Bird Count data were provided by National Audubon Society and through the generous efforts of Bird Studies Canada and countless volunteers across the Western Hemisphere.

>>>>

\pkg{sdmTMB} would not be possible without the \pkg{TMB} [@kristensen2016] and \proglang{R}-\pkg{INLA} [@rue2009; @lindgren2011; @lindgren2015] \proglang{R} packages.
\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages (as described in the DESCRIPTION and inst/COPYRIGHTS files).
Smoother support was possible thanks to \pkg{mgcv} [@wood2017a].
We thank the authors of all these packages.
We thank S. Kotwicki, M. Lindmark, M. Martin, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
Christmas Bird Count data were provided by National Audubon Society and through the generous efforts of Bird Studies Canada and countless volunteers across the Western Hemisphere.

### 2023-03-13 doc/paper-jss/pcod.Rmd
Cap #13
Copy edit

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library("ggplot2")
library("dplyr")
library("sdmTMB")
theme_set(theme_minimal())
```

@@

```{r plot-mesh, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE, fig.cap="Delaunay triangulation mesh.", fig.pos='ht', fig.align='center'}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
plot(mesh)
```

@@

```{r laplace-resids-vis}
set.seed(123)
r <- residuals(fit)
qqnorm(r)
qqline(r)
```

These residuals can suggest model fit issues even with properly specified models due to the Laplace approximation [@thygesen2017a]. We could instead use MCMC residuals, which more accurate but slower. These are currently available in the \pkg{sdmTMBextra} package at <https://github.com/pbs-assess/sdmTMBextra> and are demonstrated in the [residuals vignette](https://pbs-assess.github.io/sdmTMB/articles/).

@@

```{r plot-fix-defects, fig.cap="Predictions based on only the non-random field elements.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, exp(est_non_rf)) +
  scale_fill_viridis_c(trans = "sqrt")
```

@@

```{r plot-spatial-effects, fig.cap="Spatial random effects only.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, omega_s) +
  scale_fill_gradient2()
```

@@

```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects only.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, epsilon_st) +
  scale_fill_gradient2()
```

@@

```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index.", fig.pos='ht', fig.align='center'}
survey_grid$area <- 4 # all 2 x 2km
pred2 <- predict(
  fit,
  newdata = survey_grid, 
  return_tmb_object = TRUE
)
ind <- get_index(pred2, area = survey_grid$area, bias_correct = TRUE)

ggplot(ind, aes(year, est / 1000)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr / 1000, ymax = upr / 1000), alpha = 0.4) +
  ylab("Biomass (t)")
```

>>>>

@@

```{r packages, message=FALSE, warning=FALSE, cache=FALSE}
library("ggplot2")
library("dplyr")
library("sdmTMB")
```

```{r theme-minimal}
theme_set(theme_minimal())
```

```{r plot-mesh, fig.align='center', fig.asp=1, out.width="3in", echo=TRUE, fig.cap="Delaunay triangulation mesh. Dots represent observations of Pacific Cod.", fig.pos='ht', fig.align='center'}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
plot(mesh)
```

@@

```{r laplace-resids-vis, fig.cap="Randomized quantile residuals."}
set.seed(123)
r <- residuals(fit)
qqnorm(r)
qqline(r)
```

These residuals can suggest model fit issues even with properly specified models due to the Laplace approximation [@thygesen2017a]. We could instead use MCMC residuals, which are more accurate but slower. These are currently available in the \pkg{sdmTMBextra} package at <https://github.com/pbs-assess/sdmTMBextra> and are demonstrated in the [residuals vignette](https://pbs-assess.github.io/sdmTMB/articles/).

@@

```{r plot-fix-defects, fig.cap="Predictions based on only the non-random-field elements---a mean effect for each year.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, exp(est_non_rf)) +
  scale_fill_viridis_c(trans = "sqrt")
```

@@

```{r plot-spatial-effects, fig.cap="Spatial random effects only. This random field is constant through time.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, omega_s) +
  scale_fill_gradient2()
```

@@

```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, epsilon_st) +
  scale_fill_gradient2()
```

@@

```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index. The line represents the estimated mean and the ribbon represents a 95\\% confidence interval.", fig.pos='ht', fig.align='center'}
survey_grid$area <- 4 # all 2 x 2km
pred2 <- predict(
  fit,
  newdata = survey_grid, 
  return_tmb_object = TRUE
)
ind <- get_index(pred2, area = survey_grid$area, bias_correct = TRUE)

ggplot(ind, aes(year, est / 1000)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr / 1000, ymax = upr / 1000), alpha = 0.4) +
  ylab("Biomass (t)") + xlab("Year")
```

### 2023-03-13 doc/paper-jss/timing.Rmd
More \pkg #13
Copy edit

Here we describe the methods underlying the speed testing in Fig.\ \ref{fig:timing}.
We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Fig.\ \ref{fig:mesh-vis-timing}).
We did this by setting the `max.edge` argument, which controls the largest allowed triangle edge length.
We tested values of `max.edge` of 0.06, 0.075, 0.1, 0.15, and 0.2.
In Figure 4, we report on the x-axis the number of mesh nodes ("knots" or "vertices") that result from each of these meshes.

@@

## Illustration of generating the \textbf{INLA} mesh

@@

```{r mgcvfit-timing, echo=TRUE, eval=TRUE}
# define smooth.construct.spde.smooth.spec() and Predict.matrix.spde.smooth()
# from supplement of:
# Miller, D.L., Glennie, R. & Seaton, A.E. (2019). Understanding the Stochastic
# Partial Differential Equation approach to smoothing. Journal of Agricultural,
# Biological and Environmental Statistics.
# https://doi.org/10.1007/s13253-019-00377-z

fit_bam <- bam(
  observed ~ a1 + s(X, Y,
    bs = "spde", k = mesh$n,
    xt = list(mesh = mesh)
  ),
  data = sim_dat,
  family = gaussian(),
  method = "fREML",
  control = gam.control(scalePenalty = FALSE), 
  discrete = TRUE
)
```

>>>>

Here, we describe the methods underlying the speed testing in Fig.\ \ref{fig:timing}.
We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Fig.\ \ref{fig:mesh-vis-timing}).
We did this by setting the `max.edge` argument, which controls the largest allowed triangle edge length.
We tested values of `max.edge` of 0.06, 0.075, 0.1, 0.15, and 0.2.
In Figure 4, we report on the x-axis the number of mesh nodes ("knots" or "vertices") that result from each of these meshes.

@@

\subsection[Illustration of generating the INLA mesh]{Illustration of generating the \pkg{INLA} mesh}

@@

First define `smooth.construct.spde.smooth.spec()` and `Predict.matrix.spde.smooth()`
from the supplement of @miller2019, then:

```{r mgcvfit-timing, echo=TRUE, eval=TRUE}
fit_bam <- bam(
  observed ~ a1 + s(X, Y,
    bs = "spde", k = mesh$n,
    xt = list(mesh = mesh)
  ),
  data = sim_dat,
  family = gaussian(),
  method = "fREML",
  control = gam.control(scalePenalty = FALSE), 
  discrete = TRUE
)
```
